Computer Assisted Verbal Autopsy: Comparing Large Language Models to Physicians for Assigning Causes to 6939 Deaths in Sierra Leone from 2019-2022

### 2025-06-26 manuscript/latex/manuscript.tex
Working on docx

A total of 11,920 verbal autopsy (VA) records were obtained from the HEAL-SL study \citep{njalauniversityHealthySierraLeone2023,carshon-marshChildMaternalAdult2022}, which employed dual-coded Electronic Verbal Autopsy (EVA). Each record was independently reviewed by two randomly selected physicians, who assigned COD codes based on the International Classification of Diseases, 10th Revision (ICD-10) \citep{worldhealthorganizationICD10InternationalStatistical2011}. Agreement between physician-assigned CODs was evaluated using Central Medical Evaluation Agreement 10 (CMEA-10) codes, which group related ICD-10 codes into broader, clinically similar categories \citep{aleksandrowiczPerformanceCriteriaVerbal2014} (see Additional File 2). If both codes fell within the same CMEA-10 group, the record was considered in agreement. Disagreements entered a reconciliation phase, where each physician was shown both the assigned codes and the reasoning from the other physician. Physicians could then (1) retain their original code, (2) adopt the other physician’s code, or (3) assign a new code. Records that remained unresolved proceeded to adjudication, where a senior physician reviewed all reasoning and assignments and issued a final COD.

To ensure comparability with physician coding, only records with physician agreement were used in this study, as such cases provide higher confidence in the COD assignment \citep{barnettComparativeAccuracyDiagnosis2019,morrisFactorsAssociatedPhysician2010,hsiaoFactorsAssociatedPhysician2012}. From the original dataset, 6,942 records met this criterion. All ICD-10 codes were then standardized to CGHR-10 categories (see Additional File 1), which group causes into 19, 10, and 7 categories for adults (12–69 years), children (28 days to 11 years), and neonates (under 28 days), respectively. After excluding three records without a valid CGHR-10 category, a total of 6,939 physician-agreed records (3,826 adult, 2,636 child, and 477 neonatal) were used for model comparison and performance evaluation. Further details on data preprocessing are provided in Appendix \ref{app-methods-data}, with COD and age group distributions summarized in Tables \ref{tab-data-cod} and \ref{tab-data-age}.

@@

Additional files were used to supplement this paper:

\begin{itemize}
    \item Additional file 1: Centre for Global Health Research 10 (CGHR-10) codes. Codes grouping ICD-10 code ranges into generalized categories. (.csv)
    \item Additional file 2: Central Medical Evaluation Agreement 10 (CMEA-10) codes. ICD-10 code ranges considered in physician agreement. (.csv)
\end{itemize}

@@

The datasets supporting the conclusions of this article are included within the article (and its additional files), at \url{https://openmortality.org} (available upon request). Verbal Autopsy (VA) and narrative data by age group and survey rounds 1 and 2 available at \url{https://openmortality.org/dataset/heal-sl}. Cause of death code mappings to convert between ICD-10, WVA-2016, and CGHR-10 codes available at \url{https://openmortality.org/dataset/icd}.

@@

PJ and PB are the study Principal Investigators. ATA and RK implemented the data collection procedures. RW, and TKSN processed, documented, and prepared the data. RW, ASL, and RK ran the models. RW wrote the paper and conducted the analysis. AB and RCM provided medical domain guidance and feedback. All authors reviewed the results and contributed to the report. All authors read and approved the final manuscript.

@@

@@

\begin{figure}[H]
\centering
\includegraphics[width=.7\textwidth]{../figures/fig-methods.pdf}
\caption{Detailed study methods.}\label{fig-methods}
\end{figure}

@@

Initially, 11,920 records were collected from dual-coded EVA in the HEAL-SL study. Physicians were able to assign CODs for 11,820 of the 11,920 records, where 100 of these records could not be assigned a COD due to missing or inadequate information (e.g. low quality narrative, data loss). The 11,820 physician coded records were further filtered for records where both physicians agreed on the assigned codes (records that were not reconciled or adjudicated) resulting in 6942 physician agreed records (based on comparisons using CMEA-10 codes, see Additional File 2). The 6942 records were converted into CGHR-10 codes (see Additional File 1) that generalized ICD-10 codes into 19, 10, and 7 categories for the adult (12 to 69 years), child (28 days to 11 years), and neonatal (under 28 days) age groups. After conversion, a final total of 6939 physician agreed records (3826 adult, 2636 child, and 477 neonatal) were used for modelling and performance evaluation, where three records were removed as their ICD-10 codes did not have a matching CGHR-10 code.

>>>>

A total of 11,920 verbal autopsy (VA) records were obtained from the HEAL-SL study \citep{njalauniversityHealthySierraLeone2023,carshon-marshChildMaternalAdult2022}, which employed dual-coded Electronic Verbal Autopsy (EVA). Each record was independently reviewed by two randomly selected physicians, who assigned COD codes based on the International Classification of Diseases, 10th Revision (ICD-10) \citep{worldhealthorganizationICD10InternationalStatistical2011}. Agreement between physician-assigned CODs was evaluated using Central Medical Evaluation Agreement 10 (CMEA-10) codes, which group related ICD-10 codes into broader, clinically similar categories \citep{aleksandrowiczPerformanceCriteriaVerbal2014} (see Additional File 1). If both codes fell within the same CMEA-10 group, the record was considered in agreement. Disagreements entered a reconciliation phase, where each physician was shown both the assigned codes and the reasoning from the other physician. Physicians could then (1) retain their original code, (2) adopt the other physician’s code, or (3) assign a new code. Records that remained unresolved proceeded to adjudication, where a senior physician reviewed all reasoning and assignments and issued a final COD.

To ensure comparability with physician coding, only records with physician agreement were used in this study, as such cases provide higher confidence in the COD assignment \citep{barnettComparativeAccuracyDiagnosis2019,morrisFactorsAssociatedPhysician2010,hsiaoFactorsAssociatedPhysician2012}. From the original dataset, 6,942 records met this criterion. All ICD-10 codes were then standardized to CGHR-10 categories (see Appendix \ref{app-cghr10}), which group causes into 19, 10, and 7 categories for adults (12–69 years), children (28 days to 11 years), and neonates (under 28 days), respectively. After excluding three records without a valid CGHR-10 category, a total of 6,939 physician-agreed records (3,826 adult, 2,636 child, and 477 neonatal) were used for model comparison and performance evaluation. Further details on data preprocessing are provided in Appendix \ref{app-methods-data}, with COD and age group distributions summarized in Tables \ref{tab-data-cod} and \ref{tab-data-age}.

@@

Additional file 1 (.csv) titled "Central Medical Evaluation Agreement 10 (CMEA-10) codes" with description "ICD-10 code ranges considered in physician agreement" was used to supplement this study.

@@

The datasets supporting the conclusions of this article are included within the article (and its additional files), at \url{https://openmortality.org} (available upon request). Verbal Autopsy (VA) narrative and questionnaire data by age group and survey rounds available at \url{https://openmortality.org/dataset/heal-sl}. Cause of death code mappings to convert between ICD-10, WVA-2016, and CGHR-10 codes available at \url{https://openmortality.org/dataset/icd}.

@@

PJ and PB are the study Principal Investigators. ATA and RK implemented the data collection procedures. RW, TKSN, and CM processed, documented, and prepared the data. RW, ASL, and RK ran the models. RW wrote the paper and conducted the analysis. AB and RCM provided medical domain guidance and feedback. All authors reviewed the results and contributed to the report. All authors read and approved the final manuscript.

@@

\section{Centre for Global Health Research 10 (CGHR-10) Codes}\label{app-cghr10}

Tables \ref{tab-codes-cghr10-adult} to \ref{tab-codes-cghr10-neo} provide grouped ICD-10 ranges in each CGHR-10 code by age group.

\begin{table}[!h]
    \caption{CGHR-10 codes for adults (12-69 years).}\label{tab-codes-cghr10-adult}%
    \begin{tabular}{p{0.3\textwidth}p{0.65\textwidth}}
    \toprule
    Adult CGHR-10 Code & ICD-10 Range\\

    Acute respiratory infections & H65-H68, H70-H71, J00-J22, J32, J36, J85-J86, P23 \\ 
    Tuberculosis & A15-A16, B90, J65 \\ 
    Diarrhoeal & A00-A09 \\ 
    Unspecified infections & A17-A33, A35-A99, B00-B17, B19-B49, B55-B89, B91-B99, C46, D64, D84, G00-G09, H10, H60, I30, I32-I33, K02, K04-K05, K61, K65, K67, K81, L00-L04, L08, M00-M01, M60, M86, N10, N30, N34, N41, N49, N61, N70-N74, P35-P39, R50, R75, ZZ21 \\ 
    Malaria & B50-B54 \\ 
    Maternal conditions & A34, F53, O00-O08, O10-O16, O20-O99 \\ 
    Nutritional deficiencies & D50-D53, E00-E02, E40-E46, E50-E64, X53-X54 \\ 
    Chronic respiratory & J30-J31, J33-J35, J37-J64, J66-J84, J90-J99, R04-R06, R84, R91 \\ 
    Cancers & C00-C26, C30-C45, C47-C58, C60-C97, D00-D48, D91, N60, N62-N64, N87, R59 \\ 
    Ischemic heart & I20-I25, R55 \\ 
    Stroke & G45-G46, G81-G83, I60-I69 \\ 
    Diabetes mellitus & E10-E14 \\ 
    Other cardiovascular & I00-I03, I05-I15, I26-I28, I31, I34-I52, I70-I99, R00-R01, R03, ZZ23 \\ 
    Liver and alcohol related & B18, F10, K70-K77, R16-R18, X45, Y15, Y90-91 \\ 
    Other noncommunicable & D55-D63, D65-D83, D86, D89, E03-E07, E15-E35, E65-E68, E70-E90, F00-F09, F11-F52, F54-F99, G10-G37, G40-G41, G43-G44, G50-G80, G84-G99, H00-H06, H11-H59, H61-H62, H69, H72-H95, K00-K01, K03, K06-K14, K20-K31, K35-K38, K40-K60, K62-K64, K66, K78-K80, K82-K93, L05, L10-L99, M02-M54, M61-M85, M87-M99, N00-N08, N11-N29, N31-N33, N35-N40, N42-N48, N50-N59, N75-N86, N88-N99, Q00-Q99, R10-R15, R19-R23, R26-R27, R29-R49, R56, R63, R70-R74, R76-R77, R80-R82, R85-R87, R90, ZZ25 \\ 
    Road and transport injuries & V01-V99, Y85 \\ 
    Suicide & X60-X84 \\ 
    Other injuries & S00-S99, T00-T99, W00-W99, X00-X44, X46-X52, X55-X59, X85-X99, Y00-Y14, Y16-Y84, Y86-Y89, Y92-Y98, ZZ27 \\ 
    Ill-defined & R02, R07-R09, R25, R51-R54, R57-R58, R60-R62, R64-R69, R78-R79, R83, R89, R92-R94, R96, R98-R99 \\

\begin{table}[!htbp]
    \caption{CGHR-10 codes for children (28 days to 11 years).}\label{tab-codes-cghr10-child}%
    \begin{tabular}{p{0.25\textwidth}p{0.7\textwidth}}
    \toprule
    Child CGHR-10 Code & ICD-10 Range\\

    Pneumonia & A37, H65-H68, H70-H71, J00-J22, J32, J36, J85-J86, P23, U04 \\ 
    Diarrhoeal & A00-A09 \\ 
    Malaria & B50-B54 \\ 
    Other infections & A15-A28, A30-A36, A38-A44, A46, A48-A71, A74-A75, A77-A99, B00-B09, B15-B27, B30, B33-B49, B55-B60, B64-B83, B85-B92, B94-B97, B99, G00-G09, H10, H60, I30, I32-I33, I39-I41, J65, K02, K04-K05, K61, K65, K67, K81, L00-L04, L08, M00-M01, M60, M86, N10, N30, N34, N41, N49, N61, N70-N74, P35-P39, R50, R75, U00, Y95, ZZ11 \\ 
    Congenital anomalies & P01, P05, P07, P21, Q00-Q99 \\ 
    Epilepsy, leukaemia, and other noncommunicable & C00-C97, D01-D48, D55-D89, E03-E35, E65-E90, F00-F02, F73, G10-G99, H00-H06, H11-H59, H61-H62, H69, H72-H95, I00-I28, I31, I34-I38, I42-I99, J30-J31, J33-J35, J37-J47, J60, J64, J66-J70, J80-J82, J84, J90-J99, K00-K01, K03, K06-K60, K62-K63, K70-K80, K82-K93, L05, L10-L99, M02-M54, M61-M85, M87-M99, N00-N08, N11-N29, N31-N33, N35-N40, N42-N48, N50-N51, N60, N62-N64, N75-N99, P04, P08, P27, P51, P53-P60, P70-P72, P74-P76, P78, P80-P83, P92-P94, R00-R01, R03-R06, R11-R23, R26-R27, R29-R49, R55-R56, R59, R63, R70-R74, R76-R77, R80-R82, R84-R87, R90-R91, ZZ12-ZZ13, ZZ15 \\ 
    Injuries & S00-S99, T00-T98, V01-V99, W00-W99, X00-X52, X57-X99, Y00-Y91, Y97-Y98 \\ 
    Nutritional deficiencies & D50-D53, E00-E02, E40-E46, E50-E56, E59-E61, E63-E64, X53-X54 \\ 
    Other & D00, F03-F72, F74-F99, P00, P02-P03, P10-P15, P20, P22, P24-P26, P28-P29, P50, P52, P61, P77, P90-P91 \\ 
    Ill-defined & P96, R02, R07, R09-R10, R25, R51-R54, R57-R58, R60-R62, R64, R68-R69, R78-R79, R83, R89, R92-R99 \\

\begin{table}[!htbp]
    \caption{CGHR-10 codes for neonates (less than 28 days).}\label{tab-codes-cghr10-neo}%
    \begin{tabular}{p{0.35\textwidth}p{0.6\textwidth}}
    \toprule
    Neonate CGHR-10 Code & ICD-10 Range\\

    Prematurity \& low birthweight & D64, O60, P01, P05, P07, P22, P25-P28, P52, P61, P77, P80, P92, R04 \\ 
    Neonatal infections & A00-A09, A20-A28, A32-A35, A37-A44, A46, A48-A49, A68-A70, A74-A75, A77-A79, A81-A90, B54, B95-B96, G00-G09, H10, H60, H65-H68, H70-H71, I30, I32-I33, I39-I41, J00-J22, J32, J36, J85-J86, K65, K67, K81, L00-L04, L08, M00-M01, M60, M86, N10, N30, N34, N41, N49, N61, O85, P23, P35-P39, P58-P59, P63, U04 \\ 
    Birth asphyxia and birth trauma & G40, P00, P02-P03, P10-P15, P20-P21, P24, P29, P50-P51, P90-P91, R06, W79, Z37 \\ 
    Stillbirth & P95 \\ 
    Congenital anomalies & C76, Q00-Q99 \\ 
    Other & A15-A19, A30-A31, A36, A50-A67, A71, A80, A91-A99, B00-B09, B15-B27, B30, B33-B53, B55-B60, B64-B83, B85-B92, B94, B97, B99, C00-C75, C77-C97, D00-D48, D50-D53, D55-D63, D65-D89, E00-E35, E40-E46, E50-E56, E59-E61, E63-E90, F00-F99, G10-G39, G41-G99, H00-H06, H11-H59, H61-H62, H69, H72-H95, I00-I28, I31, I34-I38, I42-I99, J30, J31, J33-J35, J37-J47, J60, J64-J70, J80-J82, J84, J90-J99, K00-K63, K70-K80, K82-K93, L05, L10-L99, M02-M54, M61-M85, M87-M99, N00-N08, N11-N29, N31-N33, N35-N40, N42-N48, N50-N51, N60, N62-N64, N70-N99, P04, P08, P53-P57, P60,P70-P72, P74-P76, P78, P81-P83, P93-P94, R00-R01, R03-R05, R11-R23, R26-R27, R29-R36, R39-R50, R55-R56, R59, R63, R70-R77, R80-R82, R84-R87, R90-R91, S00-S99, T00-T98, U00, V01-V99, W00-W78, W80-W99, X00-X54, X57-X99, Y00-Y91, Y95, Y97-Y98 \\ 
    Ill-defined & P96, R02, R07, R09-R10, R25, R51-R54, R57-R58, R60-R62, R64, R68-R69, R78-R79, R83, R89, R92-R99 \\

@@

\begin{figure}[!h]
\centering
\includegraphics[width=.7\textwidth]{../figures/fig-methods.pdf}
\caption{Detailed study methods.}\label{fig-methods}
\end{figure}

@@

Initially, 11,920 records were collected from dual-coded EVA in the HEAL-SL study. Physicians were able to assign CODs for 11,820 of the 11,920 records, where 100 of these records could not be assigned a COD due to missing or inadequate information (e.g. low quality narrative, data loss). The 11,820 physician coded records were further filtered for records where both physicians agreed on the assigned codes (records that were not reconciled or adjudicated) resulting in 6942 physician agreed records (based on comparisons using CMEA-10 codes, see Additional File 1). The 6942 records were converted into CGHR-10 codes (see Appendix \ref{app-cghr10}) that generalized ICD-10 codes into 19, 10, and 7 categories for the adult (12 to 69 years), child (28 days to 11 years), and neonatal (under 28 days) age groups. After conversion, a final total of 6939 physician agreed records (3826 adult, 2636 child, and 477 neonatal) were used for modelling and performance evaluation, where three records were removed as their ICD-10 codes did not have a matching CGHR-10 code.
